---
title: "Exploratory Data Analysis - Project 3"
author: "Team Gryffindor"
date: "3/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(RColorBrewer)
library(ggrepel)
library(usmap)
library(lubridate)
library(reshape2)
library(curl)
library(rvest)
```

## Exploratory Data Analysis - Project 3

This document will perform exploratory data analysis on a loan dataset from Kaggle entitled the <a href="https://www.kaggle.com/datasets/wordsforthewise/lending-club">LendingClub.</a>

**<u>Questions are as follows:</u>**

*What are the most common:*

  - Job Titles
  - Loan purpose request
  - States with loan request
  
*Does loan grade or loan status have anything to do with job title?*

*How do loan amounts relate to salaries?*

### Import data

<a href="https://www.kaggle.com/wordsforthewise/lending-club?select=accepted_2007_to_2018Q4.csv.gz"> Kaggle LendingClub </a> data set was modified by extrapolating all data prior to 2018, and imported form MySQLWorkbench into R.

```{r import-data, message=FALSE}
url <- "http://data607project3.com/list-all.php"
data_html <- read_html(url)
loan_df <- html_table(data_html, fill = T)[[1]]

# preview df
glimpse(loan_df)
```

Data  has 63 columns and 686 rows.

<a href="https://www.kaggle.com/jonchan2003/lending-club-data-dictionary">Data Dictionary</a> can be referenced to specify column meanings.

### Job type

Let's look at the `emp_title` column to see what types of jobs folks had.

```{r count-emptitle, message=FALSE}
emp_count <- loan_df %>%
  count(emp_title) %>% 
  arrange(desc(n)) %>% 
  slice_max(n, n = 10) %>% 
  ggplot(aes(reorder(emp_title, n),n)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue")) +
    labs(x="", y="", subtitle = "Employe Title Count")

ggsave("../images/emp_title_count.png", 
       plot = emp_count)

emp_count
```

Some titles are the same, just have different cases and perhaps some white space. Let's clean this up.

```{r emp-title-convert-lower-trim-ws}
loan_df <- loan_df %>% 
  mutate(emp_title = trimws(tolower(emp_title))) 
```

Let's try the count again chart again, this time with color.

```{r cont-emptitle-color-cleaned}
emp_count_col <- loan_df %>%
  count(emp_title) %>% 
  arrange(desc(n)) %>% 
  slice_max(n, n = 10, with_ties = F) %>% 
  ggplot(aes(reorder(emp_title, n),n, fill = n)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(x="", y="", title = "Employe Title Count (Top 10)") +
    theme(legend.position = "none")

ggsave("../images/emp_title_count_col.png", plot = emp_count_col)

emp_count_col
```
If we filter out NA let's see this now:

```{r count-emptitle-color-cleaned-nona}
emp_count_col_na <- loan_df %>%
  filter(emp_title != "") %>% 
  count(emp_title) %>% 
  arrange(desc(n)) %>% 
  slice_max(n, n = 10, with_ties = F) %>% 
  ggplot(aes(reorder(emp_title, n),n, fill = n)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(x="", y="", title = "Employe Title Count (Top 10)") +
    theme(legend.position = "none")

ggsave("../images/emp_title_count_col_nona.png", plot = emp_count_col_na)

emp_count_col_na
```


Now we can see a very different story here.
Top 3 are:

- `NA`
- `teacher`
- `manager`

Let's look at that as a percentage:

```{r proportion-emptitle}
emp_count_col_per <- loan_df %>% 
  count(emp_title) %>% 
  mutate(p = n/sum(n)) %>% 
  slice_max(n, n = 10, with_ties = F) %>% 
  ggplot(aes(reorder(emp_title, n),p, fill = p)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(x="", y="", title = "% by Employee Title (Top 10)") +
    scale_y_continuous(labels = percent) +
    theme(legend.position = "none")

ggsave("../images/emp_title_col_per_nona.png", emp_count_col_per)

emp_count_col_per
```

So highest proportion of job titles reported by percentage are:

  - `teacher` with ~3.5%
  - `manager` with ~3.5%
  - `owner` with ~2.4%
  
### Loan Purpose

LendingClub notates the purpose of each loan.
Let's see what the most common loan purpose.

```{r loan-purpose-count}
pur_count <- loan_df %>%
  group_by(title) %>%
  filter(title != "") %>%
  summarise(count = n()) %>%
  top_n(n = 10, wt = count) %>% 
  ggplot(aes(x = reorder(title,count), y = count, fill = count)) +
    geom_bar(stat = "identity") +
    scale_fill_gradient2(high = muted("Blue"), mid = muted("red")) +
    labs(title = "Loan Purpose Count (Top 10)", x="",y="") +
    coord_flip() +
    theme(legend.position = "none")

ggsave("../images/loan_pur_count.png", plot = pur_count)

pur_count
```

We can see `Debt consolidation` is by far the highest reason for loan purpose.<br>
This is followed by `Credit card refinancing` and `Other`.

Let's look at this as a proportion.

```{r loan-purpose-percent}
pur_per <- loan_df %>%
  filter(title != "") %>%
  count(title) %>% 
  mutate(p = n/sum(n)) %>% 
  slice_max(p, n = 10) %>%
  ggplot(aes(x = reorder(title, p), y = p, fill = p)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(title = "% Loan by Purpose (Top 10)", x = "", y = "") +
    scale_y_continuous(labels = percent) +
    theme(legend.position = "none")

ggsave("../images/loan_pur_per.png", plot = pur_per)

pur_per
```

### Grade

LendingClub assigns a grade for each loan, let's look at the different grades and then sort it with `emp_title`.

```{r grades-count}
grade_count <- loan_df %>% 
  count(grade) %>% 
  ggplot(aes(reorder(grade,n), n, fill = n)) + 
    geom_bar(stat = "identity") +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    coord_flip() +
    labs(x = "Grade", y = "Count", title = "Loan Grade Count") +
    theme(legend.position = "none")

ggsave("../images/loan_grade_count.png", grade_count)

grade_count
```

Grade by proportion.

```{r grade-proportion}
grade_per <- loan_df %>% 
  count(grade) %>% 
  mutate(p = n/sum(n)) %>% 
  ggplot(aes(reorder(grade, n),p, fill = p)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(x="", y="", title = "% by Grade") +
    scale_y_continuous(labels = percent) +
    theme(legend.position = "none")

ggsave("../images/loan_grade_per.png", grade_per)

grade_per
```

`B` is the most frequent `grade`, and makes up about 30% of loans.

Let's look at how this relates to `emp_title`.

```{r top-5-emp-title}
# create df for top 5 emp title
top_5_emp_title <- loan_df %>%
  count(emp_title) %>% 
  filter(!is.na(emp_title)) %>% 
  arrange(desc(n)) %>% 
  head(5)

# compare emp title by grade
top_5_loan_grade <- loan_df %>% 
  filter(emp_title%in%top_5_emp_title$emp_title) %>% # filter top5 emp
  group_by(emp_title, grade) %>% 
  summarise(
    count = n()
  ) %>% 
  ggplot(aes(x = factor(emp_title,
      levels = as.vector(top_5_emp_title$emp_title)),
      y = count, 
      fill = grade)) + 
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_brewer(palette = "Dark2") +
    labs(x = "", y = "Percent", title = "Top 5 Jobs by Loan Grade") +
    scale_y_continuous(labels = percent) +
    guides(fill=guide_legend(title = "Grade")) +
    theme_minimal()

ggsave("../images/top_5_jobs_grade.png", plot = top_5_loan_grade)

top_5_loan_grade
```

So we can see that of the top 5 jobs, `owner` has the largest proportion of A loans, whilst `teacher` C and E.<br>
So perhaps this suggest owner's have better graded loans then teachers. 

### Loan status

Let's look distribution of loan status.

```{r loan-status}
status <- loan_df %>%
  count(loan_status) %>%
  mutate(p = n / sum(n)) %>% 
  ggplot(aes(reorder(loan_status, p), p, fill = loan_status)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "", y = "", title = "% by Loan status") +
    scale_y_continuous(labels = percent) +
    theme(legend.position = "none") +
    scale_fill_brewer(palette = "Dark2")

ggsave("../images/loan_status.png", plot = status)

status
```

Over 80% of loans are current...who are the outliers that are late?

```{r late-loan-df}
loan_df %>% 
  filter(loan_status == "Late (31-120 days)" | 
           loan_status == "Late (16-30 days") %>% 
  group_by(emp_title) %>% 
  summarise(
    count = n()
  ) %>% 
  arrange(desc(count)) %>% 
  head(10)
```
So late loan payments seem to be an anomaly and no real pattern in regard to ``emp_title`.

### Loan attributes by State

What are the most common states where loan applications are submitted?


```{r loan-app-state}
app_state <- loan_df %>%
  group_by(addr_state) %>%
  filter(addr_state != "") %>%
  summarise(count = n()) %>%
  top_n(n = 10, wt = count) %>%
  ggplot(aes(x = reorder(addr_state,count), y = count, fill = count)) +
    geom_bar(stat = "identity") +
    scale_fill_gradient2(high = muted("Blue"), mid = muted("Red")) +
    labs(title = "Loan Counts by State (Top 10)", x="State",y="") +
    coord_flip() +
    theme(legend.position = "none")

ggsave("../images/state_loan_count.png", plot = app_state)

app_state
```

We can see `CA` has the highest count, followed by `NY` and `FL`.<br>
Let's take a look at this on a map.

```{r map-accepted-loan}
# read in states csv
states<-read.csv(curl("https://raw.githubusercontent.com/brsingh7/DATA607/main/states.csv"))

# merge with loans df
loans_bs3 <- merge(loan_df,states,by.x=c("addr_state"),by.y=c("Code"),all.y=T)


state_counts <- loans_bs3 %>%
  group_by(State) %>%
  summarise(count=n(), avg_loan=mean(loan_amnt))
  state_counts$fips <- fips(state_counts$State)
  
  apps <- plot_usmap(data = state_counts, 
                     values = "count", 
                     color = "black", labels=TRUE) + 
  scale_fill_continuous(low = "white", 
                        high = muted("blue"), 
                        name = "Loan Count", 
                        label = comma) + 
  labs(title="Accepted Loan Applications by State") +
  theme(legend.position = "right")

apps$layers[[2]]$aes_params$size <- 3

ggsave("../images/state_loan_acc_map.png", plot = apps)

apps
```

Well that's interesting, but perhaps pure counts isn't the best way to look at it, let's look at average loan amount per state. 

```{r avg-loan-state}
avg_state <- plot_usmap(data = state_counts, 
                        values = "avg_loan", 
                        color = "black", labels=TRUE) + 
  scale_fill_continuous(low = "white", 
                        high = muted("red"), 
                        name = "Loan Amt ($USD)", 
                        label = comma) + 
  labs(title="Avg Loan Amount ($USD) by State") +
  theme(legend.position = c(0.9,0.05))

avg_state$layers[[2]]$aes_params$size <- 3

ggsave("../images/state_avg_loan.png", avg_state)

avg_state
```

So now we can see the states with the highest average loan amount are:

- `Idaho`
- `Missouri`.

Interesting, well now let's get some income data from <a href = "https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_income"> Wikipedia </a>, and look at loan amounts vs. salary.

```{r loan-vs-salary, warning=FALSE, message=FALSE}
# read in income data
income_2018<-read.csv(curl("https://raw.githubusercontent.com/brsingh7/DATA607/main/Project3/income_2018.csv"))

# group by state avg loan
loans_2018<- loans_bs3 %>%
  group_by(addr_state) %>%
  summarise(avg_ln=mean(funded_amnt))

# join with salary data
loan_v_salary <- left_join(loans_2018,income_2018,by=c("addr_state"="Code"))

# scatterplot
loan_v_salary_plot <- loan_v_salary %>%
  ggplot(aes(x = avg_inc_2018, y = avg_ln, color = avg_inc_2018)) +
  geom_point() +
  geom_smooth(se=F) +
  labs(title = "Mean Income vs. Mean Loan by State (2018)", 
       x="Income",y="Loan") +
  geom_text_repel(aes(label = addr_state), size = 3) +
  scale_y_continuous(label = dollar_format(prefix = "$", 
                                           scale = .001, 
                                           suffix = "k")) +
  scale_x_continuous(label = dollar_format(prefix = "$",
                                           scale = .001,
                                           suffix = "k")) +
  scale_colour_continuous(name = "Income",
                          label = dollar_format(prefix = "$",
                                                scale = .001,
                                                suffix = "k")) +
  theme(legend.position = "right")

ggsave("../images/loan_vs.sal.png")

loan_v_salary_plot
```

We can see there is not exactly a linear relatinship between average loan amount per state and average income. However we can see `Missouri (MS)` which from the previous map we could see had the highest average loan amount, also has one of the lowest incomes.<br>
Whereas states that had the highest counts of loans like `California` and `New York` had lower average loan amounts and much higher average incomes.

## Conclusion



