---
title: "Exploratory Data Analysis - Project 3"
author: "Justin Williams"
date: "3/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import-packages, message=FALSE}
library(tidyverse)
library(RMySQL)
```

## Exploratory Data Analysis - Project 3

This document will perform exploratory data analysis on a loan dataset from Kaggle entitled the LendingClub. 

### Import data from database

```{r connection-object}
#dbConnect(MySQL(),
          #user = 'data607user',
          #password = 'harry2022',
          #dbname = 'data607p3',
          #host_local = '127.0.0.1')
```

Data table LendingClub was exported from MySQLWorkbench and imported as CSV.

```{r import-csv}
loan_df <- read_delim(file = "LendingClub.csv")

# preview df
glimpse(loan_df)
```

So far interested in the following columns:

`addr_state` - The state provided by the borrower in the loan application

`avg_cur_bal` - Average current balance of all accounts

`delinq_amt` - The past-due amount owed for the accounts on which the borrower is now delinquent.

`emp_title` - The job title supplied by the Borrower when applying for the loan.*

`emp_length` - Employment length in years. Possible values are between 0 and 10 where 0 means less than one year an...

`grade` - LC assigned loan grade

`loan_amt`- The listed amount of the loan applied for by the borrower. If at some point in time, the credit depa...

`funded_amnt` - The total amount committed to that loan at that point in time.

`int_rate` - Interest Rate on the loan

`max_bal_hc` - Maximum current balance owed on all revolving accounts

`purpose` - A category provided by the borrower for the loan request.

`revol_bal` - Total credit revolving balance

`total_bal_ex_mort` - Total credit balance excluding mortgage

`zip_code` - The first 3 numbers of the zip code provided by the borrower in the loan application.

`hardship_flag` - Flags whether or not the borrower is on a hardship plan


