---
title: "Exploratory Data Analysis - Project 3"
author: "Justin Williams"
date: "3/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import-packages, message=FALSE}
library(tidyverse)
library(RMySQL)
library(scales)
library(RColorBrewer)
```

## Exploratory Data Analysis - Project 3

This document will perform exploratory data analysis on a loan dataset from Kaggle entitled the <a href="https://www.kaggle.com/datasets/wordsforthewise/lending-club">LendingClub.</a>

### Import data from database

```{r connection-object}
#dbConnect(MySQL(),
          #user = 'data607user',
          #password = 'harry2022',
          #dbname = 'data607p3',
          #host_local = '127.0.0.1')
```

Data table LendingClub was exported from MySQLWorkbench and imported as CSV.

```{r import-csv}
loan_df <- read_delim(file = "LendingClub.csv")

# preview df
glimpse(loan_df)
```

Data  has 63 columns and 686 rows.

<a href="https://www.kaggle.com/jonchan2003/lending-club-data-dictionary">Data Dictionary</a> can be referenced to specify column meanings.

So far interested in the following columns:

`addr_state` - The state provided by the borrower in the loan application

`avg_cur_bal` - Average current balance of all accounts

`delinq_amt` - The past-due amount owed for the accounts on which the borrower is now delinquent.

`emp_title` - The job title supplied by the Borrower when applying for the loan.*

`emp_length` - Employment length in years. Possible values are between 0 and 10 where 0 means less than one year an...

`grade` - LC assigned loan grade

`loan_amt`- The listed amount of the loan applied for by the borrower. If at some point in time, the credit depa...

`funded_amnt` - The total amount committed to that loan at that point in time.

`int_rate` - Interest Rate on the loan

`max_bal_hc` - Maximum current balance owed on all revolving accounts

`purpose` - A category provided by the borrower for the loan request.

`revol_bal` - Total credit revolving balance

`total_bal_ex_mort` - Total credit balance excluding mortgage

`zip_code` - The first 3 numbers of the zip code provided by the borrower in the loan application.

`hardship_flag` - Flags whether or not the borrower is on a hardship plan

### Job type

Let's look at the `emp_title` column to see what types of jobs folks with loans out had.

```{r count-emptitle}
loan_df %>%
  count(emp_title) %>% 
  arrange(desc(n)) %>% 
  slice_max(n, n = 10) %>% 
  ggplot(aes(reorder(emp_title, n),n)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue")) +
    labs(x="", y="", subtitle = "Count by Employe Title")
```
Some titles are the same, just have different cases and perhaps some white space. Let's take care of this.

```{r emp-title-convert-lower-trim-ws}
loan_df <- loan_df %>% 
  mutate(emp_title = trimws(tolower(emp_title))) 
```

Let's try the count again chart again, this time with color.

```{r}
loan_df %>%
  count(emp_title) %>% 
  arrange(desc(n)) %>% 
  slice_max(n, n = 10, with_ties = F) %>% 
  ggplot(aes(reorder(emp_title, n),n, fill = n)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(x="", y="", subtitle = "Top 10 - Count by Employe Title") +
    theme(legend.position = "none")
```
Can see a very different story here. Now top 3 are:
- `NA`
- `teacher`
- `manager`

Let's look at that as a percentage:

```{r proportion-emptitle}
loan_df %>% 
  count(emp_title) %>% 
  mutate(p = n/sum(n)) %>% 
  slice_max(n, n = 10, with_ties = F) %>% 
  ggplot(aes(reorder(emp_title, n),p, fill = p)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(high = muted("blue"), mid = muted("red")) +
    labs(x="", y="", subtitle = "Top 10 - % by Employee Title") +
    scale_y_continuous(labels = percent) +
    theme(legend.position = "none")
```
So highest proportion of job titles reported by percentage are:
  - `teacher` with ~3.5%
  - `manager` with ~3.5%
  - `owner` with ~2.4%
  
### Grade


LendingClub assigns a grade for each loan, let's look at the different grades and then sort it with `emp_title`.

```{r grades-count}
loan_df %>% 
  count(grade) %>% 
  ggplot(aes(reorder(grade,n), n)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "Grade", y = "Count", subtitle = "Count of Grade")
```
`B` is the most frequent `grade`, let's look at how this relates to `emp_title`.

```{r loan-status}
unique(loan_df$loan_status)
```


```{r emp-title-vs-grade}
loan_df %>% 
  ggplot(aes())
```



### Scatterplots

```{r}
library(GGally)
loan_df %>% 
  ggpairs(columns = 2:5)
```

